---
title: "**Assessment 4 - Part 1: Importing files, data wrangling, mathematical operations, plots and saving code on GitHub**"
author: "**Toan Hong - s225602499**"
date: "2025-10-05"
output: 
  html_document:
    toc: yes
    fig_width: 7
    fig_height: 7
theme: spacelab
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# **Introduction**

This report presents the results for Part 1 of Assessment 4, which includes ten analytical tasks completed using RStudio and GitHub. Two data files — gene_expression.tsv and growth_data.csv — are provided and can be accessed from the following repository:

🔗 https://github.com/ghazkha/Assessment4.git

The file “gene_expression.tsv” contains RNA-seq count data for three samples of interest.

  1.Read in the file, making the gene identifiers the row names. Show a table of values for the first six    genes.

  2.Make a new column which is the mean of the other columns. Show a table of values for the first six       genes.

  3.List the 10 genes with the highest mean expression.

  4.Determine the number of genes with a mean <10.

  5.Make a histogram plot of the mean values and include it into your report.

The file “growth_data.csv” contains measurements for tree circumference growing at two sites, control site and treatment site which were planted 20 years ago.

  6.Import this csv file into an R object. What are the column names?

  7.Calculate the mean and standard deviation of tree circumference at the start and end of the study at     both sites.

  8.Make a box plot of tree circumference at the start and end of the study at both sites.

  9.Calculate the mean growth over the last 10 years at each site.

  10.Use the t.test to estimate the p-value that the 10 year growth is different at the two sites.


# **Package Installation**

```{r}
install.packages("seqinr")
install.packages("R.utils")
install.packages("readr")
install.packages("dplyr")
install.packages("ggplot2")
```


# **Load required libraries**

```{r}
suppressPackageStartupMessages({
  library(seqinr) # is a package designed to process and analyse sequence data.
  library(R.utils) # general utilities like zip and unzip
  library(readr)    # for read_tsv and read_csv
  library(dplyr)    # for data wrangling
  library(ggplot2)  # for plotting
})
```

# **Download the files**

Go to the link https://github.com/ghazkha/Assessment4.git, then click on “view raw”, copy the URL from the address bar, use the download.file command in R as following

```{r}
URL1 = "https://raw.githubusercontent.com/ghazkha/Assessment4/refs/heads/main/gene_expression.tsv"
URL2 = "https://raw.githubusercontent.com/ghazkha/Assessment4/refs/heads/main/growth_data.csv"

download.file(URL1,destfile= "gene_expression.tsv")
download.file(URL2, destfile = "growth_data.csv")
```

# **Gene expression analysis**

## Question 1. Read in the file, making the gene identifiers the row names. Show a table of values for the first six genes.

```{r}

# Read the TSV file, making the first column (gene identifiers) as rownames
gene_expression <- read_tsv("gene_expression.tsv")
gene_expression <- as.data.frame(gene_expression)
rownames(gene_expression) <- gene_expression[,1]
gene_expression <- gene_expression[,-1]  # remove first column after setting rownames

```


```{r}

head(gene_expression) # Show first 6 genes

```

## Question 2. Make a new column which is the mean of the other columns. Show a table of values for the first six genes

```{r}

gene_expression$mean_expr <- rowMeans(gene_expression) # create a new column for mean of other columns

```


```{r}

head(gene_expression) # Show first 6 genes

```

## Question 3. List the 10 genes with the highest mean expression

```{r}

top10 <- head(gene_expression[order(-gene_expression$mean_expr), ], 10) # This line finds the top 10 genes with the highest mean expression values in your dataset and saves them into a new data frame called top10.

top10

```

## Question 4. Determine the number of genes with a mean <10

```{r}
num_low <- sum(gene_expression$mean_expr < 10) # This line counts how many genes have a mean expression value less than 10, and saves that number into a variable called num_low

num_low
```

## Question 5. Make a histogram plot of the mean values 

```{r}

# This creates a histogram showing the distribution of average gene expression levels (mean_expr) across all genes

ggplot(gene_expression, aes(x = mean_expr)) +
  geom_histogram(bins = 30, fill = "steelblue", color = "black") +
  theme_minimal() +
  labs(title = "Histogram of Mean Gene Expression",
       x = "Mean expression", y = "Frequency")

```

# **Growth data analysis**

## Question 6. Import this csv file into an R object. What are the column names?

```{r}

growth_data <- read_csv("growth_data.csv") # This line imports the CSV file named growth_data.csv into R and stores it as an R object (a data frame) called growth_data

```

```{r}

colnames(growth_data) # show column names

```

## Question 7. Calculate the mean and standard deviation of tree circumference at the start and end of the study at both sites

```{r}

# Calculate mean and standard deviation for 2005
mean_2005 <- mean(growth_data$Circumf_2005_cm, na.rm = TRUE)  # mean circumference in 2005, ignoring missing values
sd_2005   <- sd(growth_data$Circumf_2005_cm, na.rm = TRUE)    # standard deviation in 2005, ignoring missing values

# Calculate mean and standard deviation for 2020
mean_2020 <- mean(growth_data$Circumf_2020_cm, na.rm = TRUE)  # mean circumference in 2020, ignoring missing values
sd_2020   <- sd(growth_data$Circumf_2020_cm, na.rm = TRUE)    # standard deviation in 2020, ignoring missing values

# Print results
mean_2005; sd_2005  # display mean and SD for 2005
mean_2020; sd_2020  # display mean and SD for 2020

```

## Question 8. Make a box plot of tree circumference at the start and end of the study at both sites.

```{r}

boxplot(growth_data$Circumf_2005_cm,
        growth_data$Circumf_2020_cm,
        names = c("2005", "2020"),
        main = "Tree Circumference Growth",
        ylab = "Circumference (cm)",
        col = c("lightblue", "lightgreen"))

```

## Question 9. Calculate the mean growth over the last 10 years at each site.

```{r}

# Create a new column for growth between 2010 and 2020
growth_data$Growth_2010_2020 <- growth_data$Circumf_2020_cm - growth_data$Circumf_2010_cm

# Calculate mean growth by Site
mean_growth_by_site <- aggregate(Growth_2010_2020 ~ Site, data = growth_data, FUN = mean)

# Print results
mean_growth_by_site

```


## Question 10. Use the t.test to estimate the p-value that the 10 year growth is different at the two sites.

```{r}

t_test_result <- t.test(Growth_2010_2020 ~ Site, data = growth_data) # This line performs a t-test to compare the mean growth between two groups defined by Site in your dataset and stores the result in the variable t_test_result

t_test_result

```

